smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world evidence studies

### 2023-08-29 manuscript/manuscript.qmd
shortened words to meet journal word count

[**Word count:**]{.underline} 2,070 words/2,000 words

@@

147 words/150 words

@@

Electronic health records (EHR) are increasingly used to conduct real-world evidence (RWE) studies to complement evidence from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021]. Due to their detailed capture of clinical parameters such as vital signs, lab measurements, physician assessments, and lifestyle factors, EHRs can significantly improve the ability to control for imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022]. However, such prognostic factors are often partially observed which challenges the statistical analysis and can result in severe bias when predicting or estimating treatment effects if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].

In order to make an informed decision about the most appropriate analytic approach, it is crucial to investigate the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study, but general guidance frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms. These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently not often performed [@carroll2020; @wood2004; @harel2012].

@@

Given these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023]. The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data (e.g., missing data imputation).

@@

The workflow is generally categorized into descriptives, pattern diagnostics, and inferential diagnostics on potentially underlying missingness mechanisms. In this section, we cover the principles behind the main package functions, a selection of parameters users can specify, the returned results and how these results can be interpreted. Examples are illustrated using a synthetic dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following MCAR, MAR and MNAR mechanisms (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).

For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns represent relevant variables, i.e., exposure, outcome, fully observed covariates and the POCs. Any non-informative columns, such as patient identifiers, date columns, or zip codes should be dropped from the dataframe before calling the functions. Throughout all functions, the user has the option to specify a vector with the column name(s) of the POC(s) that should be investigated (`covar` parameter). If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.

@@

As a first step to explore the missingness in new datasets, the `smdi` package provides a few basic functions to describe and summarize missingness across all covariates. The `smdi_summarize()` function takes the dataframe as input and returns the amount and proportion of missing observations, which can also be stratified by a grouping variable (e.g., by an exposure or outcome variable). The `smdi_vis()` function returns a corresponding bar chart plot (example @fig-examples a).

@@

To quantify such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone]. The function returns an *asmd* object which displays an aggregated summary of the average or median ASMD along with a corresponding range of minimum and maximum AMSDs for each POC, respectivley. The object also returns a detailed 'Table 1' for each POC displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the POC. The function returns a plot for each POC illustrating the ASMD for each compared patient characteristic as a graphical visualization (example @fig-examples c) [@ggplot2].

@@

The group 2 diagnostics assess the ability to predict missingness based on observed covariates via the `smdi_rf()` function. This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would indicate the model is unable to discriminate missing vs. observed values based on available data; this could be due to a mechanism that is close to MCAR or one where the missingness is associated with unmeasured data (MNAR).

The function returns an object of class *rf* which generically prints an overview of the AUC value of each POC. The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio can be chosen by the user (`train_test_ratio` parameter). The *rf* object further returns a graph for each POC displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d). This metric can be valuable for interpreting and identifying strong predictors of missingness. It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the model. In case of inflated AUC values (\>0.9), the function prompts a message to the user reporting the most important predictor. If in such a scenario another POC is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to run the diagnostics for each POC independently rather than jointly.

@@

The third group of diagnostics with the `smdi_outcome()` function examines the association of the missingness indicator of the POC and the outcome under study. The function computes both a univariate model and a model adjusted for all other covariates in the dataset. In preceding simulations, we discerned distinct patterns in both univariate and adjusted associations between the missing indicator and the outcome, closely mirroring simulated missingness mechanisms [@weberpals2023]. As one would expect, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the POC. Under MAR, given that missingness can be explained by observed covariates, a spurious association in the univariate model disappeared after adjustment. If the missingness followed any MNAR mechanism, an association was observed regardless of adjustment.

Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes. Besides the the regression type (`model` parameter), a user needs to specify the column name containing the outcome using the `form_lhs` parameter (e.g., `Surv(eventtime, status)` in case of a Cox model). The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each POC.

@@

# minimal example of a smdi_diagnose() function call
smdi_diagnose(
  data = smdi_data,
  model = "cox",
  form_lhs = "Surv(eventtime, status)",
  n_cores = 3
  ) %>% 
  smdi_style_gt()
```

The function returns an object of class *smdi* containing a table with the results of all diagnostics for each specified POC and Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**). By cross-checking the resulting diagnostic parameters to expected estimates (@fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness mechanisms and thereby help elucidate if analytic approaches such as imputation analyses are viable options.

@@

# Discussion

>>>>

[**Word count:**]{.underline} 1,969 words/2,000 words

@@

145 words/150 words

@@

Electronic health records (EHR) are increasingly used to conduct real-world evidence (RWE) studies to complement evidence from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021]. Due to their detailed capture of clinical parameters, EHRs can significantly improve the ability to control for imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022]. However, such prognostic factors are often partially observed which challenges the statistical analysis and can result in severe bias when estimating treatment effects if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].

In order to make informed decisions about the most appropriate analytic approach, it is crucial to investigate the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study, but guidance frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms. These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently not often performed [@carroll2020; @wood2004; @harel2012].

@@

Given these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023]. The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data.

@@

The workflow is generally categorized into descriptives, pattern diagnostics, and inferential diagnostics on potentially underlying missingness mechanisms. In this section, we cover the principles behind the main package functions, a selection of parameters users can specify, the returned results and how these can be interpreted. Examples are illustrated using a synthetic dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following MCAR, MAR and MNAR mechanisms (more details: <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).

For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns represent relevant variables, i.e., exposure, outcome, fully observed covariates and the POCs. Any non-informative columns, such as patient identifiers, date columns, or zip codes should be dropped from the dataframe before calling the functions. Throughout all functions, users have the option to specify a vector with the column name(s) of the POC(s) that should be investigated (`covar` parameter). If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.

@@

As a first step to explore the missingness in new datasets, the `smdi` package provides a few basic functions to describe and summarize missingness across all covariates. The `smdi_summarize()` function returns the amount and proportion of missing observations, which can also be stratified by a grouping variable. The `smdi_vis()` function returns a corresponding bar chart plot (example @fig-examples a).

@@

To quantify such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone]. The function returns an *asmd* object which displays an aggregated summary of the average or median ASMD along with a corresponding range of minimum and maximum AMSDs for each POC, respectively. The object also returns detailed 'Table 1's' and plots [@ggplot2] for each POC displaying the distributions of observed covariates and resulting ASMDs between patients with and without an observed value for the POC (example @fig-examples c).

@@

Group 2 diagnostics assess the ability to predict missingness based on observed covariates via the `smdi_rf()` function. This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would indicate the model is unable to discriminate missing vs. observed values based on available data; this could be due to a mechanism that is close to MCAR or one where the missingness is associated with unmeasured data (MNAR).

The function returns an object of class *rf* which generically prints an overview of the AUC value of each POC. The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio can be specified (`train_test_ratio` parameter). The *rf* object further returns a graph for each POC displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d). This metric can be valuable for interpreting and identifying strong predictors of missingness. It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the model. In case of inflated AUC values (\>0.9), the function prompts a message to the user reporting the most important predictor. If in such a scenario another POC is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to run the diagnostics for each POC independently rather than jointly.

@@

The third group of diagnostics with the `smdi_outcome()` function examines the association of the missingness indicator of the POC and the outcome under study. The function computes both a univariate model and a model adjusted for all other covariates in the dataset. In preceding simulations, we discerned distinct patterns in both univariate and adjusted associations between the missing indicator and the outcome, closely mirroring simulated missingness mechanisms [@weberpals2023]. As expected, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the POC. Under MAR, given that missingness can be explained by observed covariates, a spurious association in the univariate model disappeared after adjustment. If the missingness followed any MNAR mechanism, an association was observed regardless of adjustment.

Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes. Besides the the regression type (`model` parameter), users need to specify the column containing the outcome using the `form_lhs` parameter (e.g., `Surv(eventtime, status)` in case of a Cox model). The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each POC.

@@

# minimal example of a smdi_diagnose() function call
smdi_diagnose(
  data = smdi_data,
  model = "cox",
  form_lhs = "Surv(eventtime, status)",
  n_cores = 3
  )
```

The function returns an object of class *smdi* containing a table with the results of all diagnostics for each specified POC and Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**). By cross-checking all resulting diagnostic parameters to expected estimates (@fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness mechanisms and thereby help elucidate if analytic approaches such as imputation analyses are viable options.

@@

-->
```
# Discussion
